Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis

### 2023-04-07 Write-ups/Manuscript_edit.tex
WNAR paper compitition submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage[]{mathpazo}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}\doublespacing
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis},
  pdfauthor={Ying Jin; Andrew Leroux},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{April 06, 2023}

@@

Biomedical investigators are often interested in predicting future
observations of subjects based on their historical data, typically
referred to as dynamic prediction. Traditionally, this type of data has
been modeled either marginally, such as generalized estimating
equations, or conditionally on specific subjects, such as mixed effect
models (\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
\protect\hyperlink{ref-liang1986}{LIANG and ZEGER 1986};
\protect\hyperlink{ref-lindstrom1990}{Lindstrom and Bates 1990};
\protect\hyperlink{ref-davidian2003}{{``Nonlinear models for repeated
measurement data''} 2003}). For subject-specific models, predictions are
made based on the correlation between repeated measures from the same
subject and covariates that can be either fixed or varying. Software and
methods for dynamic prediction have been developed for Gaussian
{[}leroux2016; Shang (\protect\hyperlink{ref-shang2017}{2017}); Goldberg
et al. (\protect\hyperlink{ref-goldberg2014}{2014}){]}, non-Gaussian
data (\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}), and
time-to-event outcomes (\protect\hyperlink{ref-jmbook}{Rizopoulos 2012};
\protect\hyperlink{ref-suresh2017}{Suresh et al. 2017}). However, with
the notable exception of Leroux et al.
(\protect\hyperlink{ref-leroux2016}{2018a}), existing methods are either
limited in their flexibility (both fixed and random effect structure,
e.g.~random slope and intercept), computationally unfeasible for large
datasets, or both. These issues are particularly acute for data which
are measured densely, as is often the case with functional data, or when
data are non-Gaussian.

While methods with accompanying software for modeling complex functional
data with non-linear fixed and random effects exist for both Gaussian
and non-Gaussian data, software is largely limited to the refund package
in the statistical software R
(\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014};
\protect\hyperlink{ref-wood2014}{Wood and Scheipl 2014};
\protect\hyperlink{ref-refundpkg}{\textbf{refundpkg?}}). However, these
methods and software implementations do not currently allow for dynamic
prediction. In addition to the methods implemented in the refund
package, other work has been done on the fitting of functional
regression models to non-Gaussian functional data. For example, Chen et
al. (\protect\hyperlink{ref-chen2013}{2013}) proposed approaches to fit
marginal functional models that is compatible to multi-level,
generalized outcomes. Goldsmith et al.
(\protect\hyperlink{ref-goldsmith2015}{2015}) established a model
framework that takes into account the fixed effect of time-invariant
covariates, with parameters estimated with Bayesian method in
\emph{Stan}. Gertheiss et al.
(\protect\hyperlink{ref-gertheiss2016}{2016}) identified bias introduced
by directly applying FPCA methods to generalized functions, and proposed
to address this problem using a two-stage joint estimation strategy.
Linde (\protect\hyperlink{ref-linde2019}{2009}) used an adapted Bayesian
variational algorithm for FPCA of binary and count data. In terms of
implementation, Wrobel et al. (\protect\hyperlink{ref-wrobel2019}{2019})
proposed a fast, efficient way to fit GFPCA on binary data using EM
algorithm, accompanied by the an open source R package \emph{registr}.

Existing research on dynamic prediction with functional data has focused
primarily on continuous/Gaussian outcomes, usually modelling
subject-specific random effects with FPCA
(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
predict missing observations in partially observed function tracks, and
Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
similar goals using Markov Chains. While methods mentioned above used
only partial observations for prediction with an intercept-only model,
Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018a}) proposed
Functional Concurrent Regression (FCR) framework which can incorporate
the effect of subject-specific predictors. Although methodological
development exists, flexible software for implementing these methods is
still relatively sparse with the exception of the fcr
(\protect\hyperlink{ref-fcrpkg}{\textbf{fcrpkg?}}) package accompanying
(\protect\hyperlink{ref-leroux2016}{Leroux et al. 2018a}).

@@

The observed data for a single subject \(i\) is (\(t\), \(Y_i(t)\)),
where \(t\) consists of dense, discrete points along the domain of data
collection. In practice, \(t\) is usually a time index along the
duration of study. We hereafter refer to the range of \(t\) as the
``functional domain''. For simplicity, we further assume that outcomes
are collected on a regular grid over the functional domain, meaning that
\(t\) are evenly distributed and does not change across subjects.

\(Y_i(t)\) is the non-Gaussian outcome observed at \(t\). We assume that
the outcome \(Y_i(t)\) can be characterized by a latent continuous
function \(\eta_i(t)\). That is, at a specific t, \(Y_i(t)\) follows a
exponential family distribution such that:

@@

Estimation of \(b_i(t)\) involves estimating both
\(\{\phi_k(t): t \in \cal{T}, 1 \leq k \leq K\}\) and
\(\{\xi_{ik}: 1 \leq i \leq N, 1 \leq k \leq K\}\). For Gaussian data
this may be done by directly estimating the covariance operator
\(\Sigma_b\) using either method of moments (if the data are relatively
small) or using faster methods such as the FACE method {[}reference -
FACE{]}. However, for non-Gaussian data, direct estimation is not
possible. Instead, one may choose to estimate \(b_i(t)\) first using a
flexible spline basis (e.g.~B-splines) with dimension \(K^b\) such that
\(K^B > K\), assuming an unstructured correlation between
subject-specific spline coefficients, then orthogonalizing the resulting
estimates via eigendecomposition or singular value decomposition. This
general approach forms the basis for estimation for several GFPCA
methods {[}references - goldsmith, julia's EM algorithm{]}. However,
this approach results in rapidly escalating computational burden as both
the number of subjects, \(N\), and the density of the observed
functional predictor increases. Note that if the covariance operator
\(\Sigma_b\) is relatively complex, then \(K\) may be large and not well
approximated using simple basis functions, \(\phi_k\) (e.g.~random
intercept and slope).

Beyond the computational burden associated with fitting the model of
interest, for dynamic prediction one must then estimate the scores for a
new subject \(u\) not used in model fitting. Typically this takes the
form of the conditional expectation of the scores
\(E[\xi_{uk}|\bf{\Theta}, \bf{y}_{u}]\) where \(\bf{\Theta}\) represents
the set of parameters estimated by the model (i.e.~parameters associated
with \(\beta_0(t), \Sigma_b\)) and \(\bf{y}_u\) are the non-Gaussian
data for the new subject. Estimation of the random effects for a new
subject is non-trivial and not implemented for even very simple models
in most random effects software packages such as lme4 {[}reference{]}
and nlme {[}reference{]}. Our proposed method bypasses this problem by
using local generalized linear mixed models to estimate the covariance
operator \(\Sigma_b\) directly on the latent scale, then using the
closed form solution for the best linear unbiased predictor of Gaussian
random effects. Estimation is made fast for even very large data through
the use of the FACE methodology for FPCA on the latent estimates
obtained from local models. Because the local models are very simple
GLMMs, estimation is also fast for the local models.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Bin the observed outcomes in to small, non-overlapping, equal length
  intervals. We hereafter index the bins by their midpoints \(s\). The
  number of observations from the same subject in bin \(s\) is \(n_s\),
  and \(t_{sj}\) refers to the jth observation point in bin \(s\).
\item
  Fit a local Generalized Mixed Model at every bin. Specifically, at bin
  \(s\), we fit \(g[E(Y_i(t_{sj}))] = \beta_0(s)+b_i(s)\),
  \(j\in\{1, ..., n_s\}\). From this series of models we can estimate
  the individual latent functions at every bin:
  \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
\item
  Fit FPCA on the estimated latent functions:
  \(\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)\),
  and obtain estimates of basis functions
  \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
  \(\hat{\lambda}_1...\hat{\lambda}_k\), mean function \(\hat{f}_0(s)\)
  and residual variance \(\hat{\sigma}^2\).
\end{enumerate}

\textbf{Ying, include the formulas for your conditional expectation here
to obtain random effects estimates }

\textbf{Dynamic prediction} Now assume we have a new observations with
partially observed outcome \(Y_{new}(t)\), \(t\leq t_m\). The maximum
observed point is \(t_m\), which is a point along the functional domain
and belongs to bin \(s_m\) on the binned grid from step 1 above. With
components extracted from the FPCA model in step 3, we will be able to
make predictions on the future, unobserved bins. Specifically, with the
partially observed binary function \(Y_{new}(t)\), we will be able to
calculate the maximum likelihood estimate (MLE) of the subject-specific
PC scores \(\hat{\xi}_{ik}\). Then the value of latent functions at
unobserved points can be estimated as
\(\hat{\eta_i}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)\)

@@

In this simulation, we set \(f_0(t) = 0\). \(\xi_{ik}\) are mutually
independent normal random variables \(\xi_{ik}\sim N(0, \lambda_k)\).
Here we set the values of \(\lambda_k\) to be \(0.5^{k-1}\),
\(k \in (1,..., 4)\). In addition, for simplicity of presentation, we
generate data on a regular grid, which means observations points are
equally distributed across \((0, 1000]\) and are the same for all
subjects.

@@

As a reference method, we compare our method to Generalized Linear Mixed
Models using Adaptive Gaussian Quadrature (GLMMadaptive). This is one of
the fastest existing method developed for dynamic prediction of repeated
generalized outcomes. Just like many mixed models, this method is very
limited in terms of flexibility if one wishes to obtain estimates in a
reasonable time frame. For example, the model used for prediction of our
simulated datset would simply be an linear model with one covariate
indicating observation time:
\(g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\). While the flexibility
of this mixed model can be increased using spline functions, the
dimension of spline basis is also restricted by computational ability,
and is unfeasible to implement under the scale of our simulated data or
the complexity of our proposed method.

@@

\begin{figure}
\centering
\includegraphics{Manuscript_edit_files/figure-latex/Figure-1.pdf}
\caption{Predicted track of four randomly selected subjects from one
simulated dataset. The grey solid line indicates the true latent
continuous function. The dashed lines indicated predicted latent
function tracks, and color indicates different observation time.}
\end{figure}

@@

In this section, we illustrate the performance of our proposed method
using data from a randomized trial on weight loss at University of
Colorado Anschutz Medical Campus
(\protect\hyperlink{ref-bothwell2022}{Bothwell et al. 2022};
\protect\hyperlink{ref-ostendorf2022}{Ostendorf et al. 2022}). The study
originally aimed to study the effect of dietary strategies on weight
loss, and participants were given a smart scale for daily at-home
weighing. This record was then scaled into a binary indicator of
participant adherence to the daily-weigh-in procedure, where 1 indicates
participants followed the requirement and 0 otherwise. In this section,
we will use this series of daily indicators as the binary functional
outcome, and use the proposed method to dynamically predict participant
adherence on future days given historical track.

The dataset includes 55 participants aged 22-56 at baseline. Adherence
to daily weigh-in is available along 400 consecutive days for all
participants. In other words, for each individual, 400 binary
measurements were taken on a regular grid along the functional domain
(number of days into the study). We will fit both fGFPCA and
GLMMadaptive models and compared their performance on this data. For
fGFPCA, every 10 observations are binned together, resulting in 40 bins
for final model fitting. Linear interpolation was used for extension
from the binned grid back to the un-binned grid. Since the latent
continuous function tracks are not known in this case, prediction
performance on evaluated with AUC alone.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage[]{mathpazo}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}\doublespacing
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{}
\date{\vspace{-2.5em}}

@@

Biomedical investigators are often interested in predicting future
observations of subjects based on their historical data, typically
referred to as dynamic prediction. Traditionally, this type of data has
been modeled either marginally, such as generalized estimating
equations, or conditionally on specific subjects, such as mixed effect
models (\protect\hyperlink{ref-davidian2003}{Davidian and Giltinan
2003}; \protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
\protect\hyperlink{ref-liang1986}{Liang and Zeger 1986};
\protect\hyperlink{ref-lindstrom1990}{Lindstrom and Bates 1990}). For
subject-specific models, predictions are made based on the correlation
between repeated measures from the same subject and covariates that can
be either fixed or varying. Software and methods for dynamic prediction
have been developed for Gaussian
(\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
\protect\hyperlink{ref-leroux2016}{Leroux et al. 2018a};
\protect\hyperlink{ref-shang2017}{Shang 2017}), non-Gaussian data
(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}), and
time-to-event outcomes (\protect\hyperlink{ref-jmbook}{Rizopoulos 2012};
\protect\hyperlink{ref-suresh2017}{Suresh et al. 2017}). However, with
the notable exception of Leroux et al.
(\protect\hyperlink{ref-leroux2016}{2018a}), existing methods are either
limited in their flexibility (both fixed and random effect structure,
e.g.~random slope and intercept), computationally unfeasible for large
datasets, or both. These issues are particularly acute for data which
are measured densely, as is often the case with functional data, or when
data are non-Gaussian.

While methods with accompanying software for modeling complex functional
data with non-linear fixed and random effects exist for both Gaussian
and non-Gaussian data, software is largely limited to the \emph{refund}
package in the statistical software R
(\protect\hyperlink{ref-refundpkg}{Goldsmith et al. 2022};
\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014};
\protect\hyperlink{ref-wood2014}{Wood and Scheipl 2014}). However, these
methods and software implementations do not currently allow for dynamic
prediction. In addition to the methods implemented in the refund
package, other work has been done on the fitting of functional
regression models to non-Gaussian functional data. For example, Chen et
al. (\protect\hyperlink{ref-chen2013}{2013}) proposed approaches to fit
marginal functional models that is compatible to multi-level,
generalized outcomes. Goldsmith et al.
(\protect\hyperlink{ref-goldsmith2015}{2015}) established a model
framework that takes into account the fixed effect of time-invariant
covariates, with parameters estimated with Bayesian method in
\emph{Stan}. Gertheiss et al.
(\protect\hyperlink{ref-gertheiss2016}{2016}) identified bias introduced
by directly applying functional principal analysis (FPCA) methods to
generalized functions, and proposed to address this problem using a
two-stage joint estimation strategy. Linde
(\protect\hyperlink{ref-linde2019}{2009}) used an adapted Bayesian
variational algorithm for FPCA of binary and count data. In terms of
implementation, Wrobel et al. (\protect\hyperlink{ref-wrobel2019}{2019})
proposed a fast, efficient way to fit generalized FPCA (GFPCA) on binary
data using EM algorithm, accompanied by the an open source R package
\emph{registr}.

Existing research on dynamic prediction with functional data has focused
primarily on continuous/Gaussian outcomes, usually modelling
subject-specific random effects with FPCA
(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
predict missing observations in partially observed function tracks, and
Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
similar goals using Markov Chains. While methods mentioned above used
only partial observations for prediction with an intercept-only model,
Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018a}) proposed
Functional Concurrent Regression (FCR) framework which can incorporate
the effect of subject-specific predictors. Although methodological
development exists, flexible software for implementing these methods is
still relatively sparse with the exception of the fcr
(\protect\hyperlink{ref-fcrpkg}{Leroux et al. 2018b}) package
accompanying (\protect\hyperlink{ref-leroux2016}{Leroux et al. 2018a}).

@@

The observed data for a single subject \(i\) is (\(t\), \(Y_i(t)\)),
where \(t\) consists of dense, discrete points along the domain of data
collection. In practice, \(t\) is usually a time index along the
duration of study. We hereafter refer to the range of \(t\) as the
``functional domain'' denoted by \(T\). For simplicity, we further
assume that outcomes are collected on a regular grid over the functional
domain, meaning that \(t\) are evenly distributed and does not change
across subjects.

\(Y_i(t)\) is the non-Gaussian outcome observed at \(t\). We assume that
the outcome \(Y_i(t)\) can be characterized by a latent continuous
function \(\eta_i(t)\). That is, at a specific t, \(Y_i(t)\) follows an
exponential family distribution such that:

@@

Estimation of \(b_i(t)\) involves estimating both
\(\{\phi_k(t): t \in T, , 1 \leq k \leq K\}\) and
\(\{\xi_{ik}: 1 \leq i \leq N, 1 \leq k \leq K\}\). For Gaussian data
this may be done by directly estimating the covariance operator
\(\Sigma_b\) using either method of moments (if the data are relatively
small) or faster methods such as fast covariance estimation (FACE)
(\protect\hyperlink{ref-face}{Xiao et al. 2016}). However, for
non-Gaussian data, direct estimation is not possible. Instead, one may
choose to estimate \(b_i(t)\) first using a flexible spline basis
(e.g.~B-splines) with dimension \(K^B\) such that \(K^B > K\), assuming
an unstructured correlation between subject-specific spline
coefficients, then orthogonalize the resulting estimates via
eigendecomposition or singular value decomposition. This general
approach forms the basis for estimation for several GFPCA methods
(\protect\hyperlink{ref-goldsmith2015}{Goldsmith et al. 2015};
\protect\hyperlink{ref-wrobel2019}{Wrobel et al. 2019}). However, this
approach results in rapidly escalating computational burden as both the
number of subjects, \(N\), and the density of the observed functional
predictor increases. Note that if the covariance operator \(\Sigma_b\)
is relatively complex, then \(K\) may be large and not well approximated
using simple basis functions, \(\phi_k\) (e.g.~random intercept and
slope).

Beyond the computational burden associated with fitting the model of
interest, for dynamic prediction one must then estimate the scores for a
new subject \(u\) not used in model fitting. Typically for non-Gaussian
data, this takes the form of the maximum likelihood estimator (MLE) of
the scores
\(\hat{\xi}_{uk} = \underset{\xi_{uk}}{\mathrm{argmax}} L(\xi_{uk}|\boldsymbol{\Theta}, \boldsymbol{Y}_{u})\),
where \(\bf{\Theta}\) represents the set of parameters estimated by the
model (i.e.~parameters associated with \(\beta_0(t), \Sigma_b\)) and
\(\boldsymbol{Y}_u\) are the non-Gaussian data for the new subject.
Estimation of the random effects for a new subject is non-trivial and
not implemented for even very simple models in most random effects
software packages such as \emph{lme4}
(\protect\hyperlink{ref-lme4}{Bates et al. 2015}) and \emph{nlme}
(\protect\hyperlink{ref-nlme}{Pinheiro et al. 2022}). Our proposed
method bypasses this problem by using local generalized linear mixed
models (GLMMs) to estimate the covariance operator \(\Sigma_b\) directly
on the latent scale, then using the MLE of Gaussian random effects.
Estimation is made fast for even very large data through the use of the
FACE methodology for FPCA on the latent estimates obtained from local
models. Because the local models are very simple GLMMs, estimation is
also fast for the local models.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Bin the observed outcomes in to small, non-overlapping, equal length
  intervals. We hereafter index the bins by their midpoints \(s\). The
  number of observations from the same subject in bin \(s\) is \(n_s\),
  and \(t_{sj}\) refers to the jth observation point in bin \(s\).
\item
  Fit a local Generalized Mixed Model at every bin. Specifically, at bin
  \(s\), we fit \(g[E(Y_i(t_{sj}))] = \beta_0(s)+b_i(s)\),
  \(j\in\{1, ..., n_s\}\). From this series of models we can estimate
  the individual latent functions at every bin:
  \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
\item
  Fit FPCA on the estimated latent functions:
  \(\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)\),
  and obtain \(\hat{\boldsymbol{\Theta}}\) which includes the estimates
  of basis functions
  \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
  \(\hat{\lambda}_1...\hat{\lambda}_k\), mean function \(\hat{f}_0(s)\)
  and residual variance \(\hat{\sigma}^2\).
\end{enumerate}

\textbf{Dynamic prediction} Now assume we have a new observations \(u\)
with partially observed outcome
\(\boldsymbol{Y}_u = \{Y_u(t), t\leq t_m\}\). The maximum observed point
is \(t_m\), which is a point along the functional domain and belongs to
bin \(s_m\) on the binned grid from step 1 above. With components
extracted from the FPCA model in step 3, we will be able to make
predictions on the future, unobserved bins. Specifically, with the
partially observed binary function \(Y_u(t)\), we will be able to
calculate the MLE of the subject-specific PC scores
\(\hat{\boldsymbol{\xi}}_u= \{\hat{\xi}_{u1},...,\hat{\xi}_{uK}\}\) that
maximizes its posterior likelihood:

\[\hat{\boldsymbol{\xi}}_u = \underset{\boldsymbol{\xi}_u}{\mathrm{argmin}} \ L(\boldsymbol{\xi}_u|\boldsymbol{Y}_u, 
\hat{\boldsymbol{\Theta}})= \frac{p(\boldsymbol{Y}_u|\boldsymbol{\xi}_u, \hat{\boldsymbol{\Theta}})p(\boldsymbol{\xi}_u|\hat{\boldsymbol{\Theta}})}{\int p(\boldsymbol{Y}_u|\boldsymbol{\xi}_u,\hat{\boldsymbol{\Theta}})p(\boldsymbol{\xi}_u|\hat{\boldsymbol{\Theta}})d\boldsymbol{\xi}_u}\]

Then the value of latent functions at unobserved points can be estimated
as
\(\hat{\eta_u}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{uk}{\phi}_k(s)\).
In fact, since the denominator in the formula above is constant with
respect to \(\boldsymbol{\xi}_u\), we only need to maximum the
numerator. Since \(\boldsymbol{Y}_u|\boldsymbol{\xi}_u\) follows an
exponential family distribution and \(\boldsymbol{\xi}_u\) follows
multivariate Guassian distribution, it could be numerically maximized
efficiently.

@@

In this simulation, we set \(f_0(t) = 0\). \(\xi_{ik}\) are mutually
independent normal random variables \(\xi_{ik}\sim N(0, \lambda_k)\).
Here we set the values of \(\lambda_k\) to be \(0.5^{k-1}\),
\(k \in \{1,..., 4\}\). In addition, for simplicity of presentation, we
generate data on a regular grid, which means observations points are
equally distributed across \((0, 1000]\) and are the same for all
subjects.

@@

As a reference method, we compare our method to Generalized Linear Mixed
Models using Adaptive Gaussian Quadrature
(GLMMadaptive)(\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}).
This is one of the fastest existing method developed for dynamic
prediction of repeated generalized outcomes. Just like many mixed
models, this method is very limited in terms of flexibility if one
wishes to obtain estimates in a reasonable time frame. For example, the
model used for prediction of our simulated datasets would simply be an
linear model with one covariate indicating observation time:
\(g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\). While the flexibility
of this mixed model can be increased using spline functions, the
dimension of spline basis is also restricted by computational ability,
and is unfeasible to implement under the scale of our simulated data or
the complexity of our proposed method.

@@

\begin{figure}
\centering
\includegraphics{Manuscript_edit_files/figure-latex/Figure-1.pdf}
\caption{Predicted tracks of four randomly selected subjects from one
simulated dataset. The grey solid line indicates the true latent
continuous function. The dashed lines indicate predicted latent function
tracks, and color indicates different observation time.}
\end{figure}

@@

In this section, we illustrate the performance of our proposed method
using data from a randomized trial on weight loss at University of
Colorado Anschutz Medical Campus
(\protect\hyperlink{ref-bothwell2022}{Bothwell et al. 2022};
\protect\hyperlink{ref-ostendorf2022}{Ostendorf et al. 2022}). The trial
originally aimed to study the effect of dietary strategies on weight
loss, and participants were given a smart scale for daily at-home
weighing. This record was then scaled into a binary indicator of
participant adherence to the daily-weigh-in procedure, where 1 indicates
participants followed the requirement and 0 otherwise. In this section,
we will use this series of daily indicators as the binary functional
outcome, and use the proposed method to dynamically predict participant
adherence on future days given historical track.

The dataset includes 55 participants aged 22-56 at baseline. Adherence
to daily weigh-in is available along 400 consecutive days for all
participants. In other words, for each individual, 400 binary
measurements were taken on a regular grid along the functional domain
(number of days into the study). We will fit both fGFPCA and
GLMMadaptive models and compare their performance on this data. For
fGFPCA, every 10 observations are binned together, resulting in 40 bins
for final model fitting. Linear interpolation was used for extension
from the binned grid back to the un-binned grid. Since the latent
continuous function tracks are not known in this case, prediction
performance is evaluated with AUC alone.
